\thetitle

### 2021-05-20 writing/rational-effort.tex
tighten prose, get fig. 3 on same page as first reference

Comparing the effort of different modes of the rational programmer can
illuminate the comparitive effectiveness of the three gradual typing
systems. If, for instance, both
Natural blame and Transient first blame trails are successful for the same scenario, the two modes of
the rational programmer can compete to see which one debugs the
scenario with less effort. In general, if the effort distribution for a mode of
the rational programmer has a shorter tail and more volume around smaller values
compared to the effort distribution of another mode, then the first mode is
probably the more effective of the two.

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance.
In particular, the effort distribution for a mode of the rational programmer can be compared with that of
the random mode that ignores error information entirely and instead selects which
module to type randomly.  Since each mutant has a finite number of
modules, random mode trails are always successful. However, the random mode's effort
distribution should be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

>>>>

Comparing the effort of different modes of the rational programmer can
illuminate the comparitive effectiveness of the three gradual typing
systems. If, for instance, both
Natural blame and Transient-first blame trails are successful for the same scenario, the two modes of
the rational programmer can compete to see which one debugs the
scenario with less effort. In general, if the effort distribution for a mode of
the rational programmer has a shorter tail and more volume around smaller values
compared to the effort distribution of another mode, then the first mode is
probably the more effective of the two.

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance.
In particular, the effort distribution for one mode can be compared with that of
the random mode that ignores error information entirely and instead selects which
module to type randomly.  Since each mutant has a finite number of
modules, random mode trails are always successful. However, the random mode's effort
distribution should be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

### 2021-05-20 writing/rational-experiment.tex
Typo fixes
tighten prose, get fig. 3 on same page as first reference

\begin{figure}[ht] \footnotesize
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
\end{tabular}}
  \label{fig:experiment-outline}
\end{figure}

The above table summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example,
experimental question $Q_1$ asks whether blame is valuable for Natural and the
experiment uses the Natural blame and exception trails to answer it, so $Q_1$
shows up in the cells for Natural blame and Natural exceptions.

@@

In detail, the answer to $Q_1$ demands a comparison of the success of the Natural
blame and Natural exception trails for all debugging scenarios.  The
first step is to construct each mutant's scenario lattice and identify their
debugging scenarios.  The test bed then extends the trails that start
from those roots according to the Natural-blame programmer.  If no scenarios can
be added to the trail, the test bed checks whether the last scenario
of the trail type-checks or not. If it does, the
Natural-blame trail is successful; otherwise it is failing. 
Figure~\ref{fig:process} summarizes this experimental process for
one mode of the rational programmer and connects it with the mutations from
section~\ref{sec:mutate}.
The process is repeated for the same roots with the other mode, Natural
exceptions.
After completing, the test bed
reports the success/failure results of the trails to determine the proportion of
scenarios where Natural blame is more useful than Natural exceptions.  Question
$Q_1$ has a positive answer if a root exists where the above is true because it
is evidence that there is at least one interesting scenario that the rational
programmer manages to debug because of blame information.  The process is analogous
for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.

@@

For $Q_*$, the process is a bit more involved. Answering this question calls for
a comparison of the percentage of scenarios where one mode is more useful than
the other and the inverse.  For instance, deciding whether blame for Natural is
more useful than first blame for Transient requires comparing the percentage
of scenarios where Natural blame is more successful than Transient
first blame with the percentage where Transient first
blame is more successful than Natural blame. Repeating the whole process
for every pair of modes produces a
complete picture of the comparative usefulness of blame between all of the modes.

>>>>

\begin{wrapfigure}{r}{.450\textwidth}
% \begin{figure}[ht]
\footnotesize
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
\end{tabular}}
  \label{fig:experiment-outline}
\end{wrapfigure}

The nearby table summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example,
experimental question $Q_1$ asks whether blame is valuable for Natural and the
experiment uses the Natural blame and exception trails to answer it, so $Q_1$
shows up in the cells for Natural blame and Natural exceptions.

@@

In detail, the answer to $Q_1$ demands a comparison of the success of the Natural
blame and Natural exception trails for all debugging scenarios.  The
first step is to construct each mutant's scenario lattice and identify their
debugging scenarios.  The test bed then extends the trails that start
from those roots according to the Natural-blame programmer.  If no scenarios can
be added to the trail, the test bed checks whether the last scenario
of the trail type-checks or not. If it does not, the
Natural-blame trail is successful; otherwise it is failing. 
Figure~\ref{fig:process} summarizes this experimental process for
one mode of the rational programmer and connects it with the mutations from
section~\ref{sec:mutate}.
The process is repeated for the same roots with the Natural-exceptions mode.
After completing, the test bed
reports the success/failure results of the trails to determine the proportion of
scenarios where Natural-blame is more useful than Natural-exceptions.  Question
$Q_1$ has a positive answer if a root exists where the above is true because it
is evidence that there is at least one interesting scenario that the rational
programmer manages to debug because of blame information.  The process is analogous
for $Q_2$ and $Q_3$, using the respective modes. % of the rational programmer.

@@

For $Q_*$, the process is a bit more involved. Answering this question calls for
a comparison of the percentage of scenarios where one mode is more useful than
the other and the inverse.  For instance, deciding whether blame for Natural is
more useful than Transient-first requires comparing the percentage
of scenarios where the first is more successful than the second with
the percentage where the second is more successful than the first. Repeating the whole process
for every pair of modes produces a
complete picture of the comparative usefulness of blame. % between all of the modes.

### 2021-05-20 writing/rational-natural.tex
tex problems fixed
tighten prose, get fig. 3 on same page as first reference

Here is a rigorous definition. 
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and
\[ \conf_{i+1} \setminus \conf_i = %
\left\{ %
\begin{tabular}{ll} %
$\{\blame{\system, \conf_i}\}$      & if $\conf_i$ produces blame \\ %
$\{\exception{\system, \conf_i}\}$  & otherwise \\ %
\end{tabular} %
\]
where $\blame{\system, \conf}$ denotes the module (of $\system$) that $\conf$ blames,
and $\exception{\system, \conf}$ denotes the first untyped location in the stacktrace produced by $\conf$.
\end{quote}

This definition rests on the two options for $\conf_{i+1} \setminus \conf_i$.
The first part reflects the rational programmer's use of blame to extend the
trail. However, blame may not be available in some scenarios. The second part
accounts for those scenarios that produce errors from the runtime system before
any impedance mismatch can be detected. For example, the fully untyped
configuration can only produce such an error (e.g. from {\tt +} receiving a
string). Exceptions from the runtime do not carry blame, so
the rational programmer proceeds using the accompanying stacktrace instead.

When the buggy untyped module of a program is replaced by the typed counterpart,
the type checker fails because this module causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenario successfully pinpoints the location of
the bug. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
an untyped module. Thus the rational programmer gets
no hints from the gradual type system on how to continue the search for the bug.

@@

To account for this situation, a new mode of the Natural rational
programmer follows a migration process based on entirely on exceptions.
\begin{quote}
\it A {\em Natural exception trail\/} is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$.
\end{quote}

>>>>

\mf{Technically a Natural blame trail could be a Natural exception trail. Hmph? }

Here is a rigorous definition. 
\begin{quote}
\it A \emph{Natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and
\[ \conf_{i+1} \setminus \conf_i = %
\left\{ %
\begin{tabular}{ll} %
$\{\blame{\system, \conf_i}\}$      & if $\conf_i$ produces blame \\ %
$\{\exception{\system, \conf_i}\}$  & otherwise \\ %
\end{tabular} \right. %
\]
where $\blame{\system, \conf}$ denotes the module (of $\system$) that $\conf$ blames,
and $\exception{\system, \conf}$ denotes the first untyped location in the stacktrace produced by $\conf$.
\end{quote}

This definition rests on the two options for $\conf_{i+1} \setminus \conf_i$.
The first part reflects the rational programmer's use of blame to extend the
trail. However, blame may not be available in some scenarios. The second part
accounts for those scenarios that produce errors from the runtime system before
any impedance mismatch can be detected. For example, the fully untyped
configuration can only produce such an error, e.g. from {\tt +} receiving a
string. Exceptions from the runtime do not carry blame, so
the rational programmer proceeds using the accompanying stacktrace instead.

When the buggy untyped module of a program is replaced by the typed counterpart,
the type checker fails because this module causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenario successfully pinpoints the location of
the bug. 
\begin{quote}
\it A Natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff (the program for) its last scenario $\conf_n$ does not type check.  A Natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff (the program for) $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is, failing Natural blame trails are those that end in a scenario that does not reveal the bug statically, yet also does not blame
an untyped module. Thus the rational programmer has
no further hints on how to continue the search for the bug.

@@

To account for this situation, a new mode of the Natural rational
programmer follows a migration process based entirely on exceptions.
\begin{quote}
\it A {\em Natural exception trail\/} is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\exception{\system, \conf_i}\}$.
\end{quote}

### 2021-05-20 writing/rational-stuff.tex
tighten prose, get fig. 3 on same page as first reference

\citet{tfgnvf-popl-2016} describe the set of all possible type migrations with a
lattice. Formally, a program $\system$ is a set of modules $\set{\component}$.
Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$
that comes with type annotations. These configurations of $\system$ are ordered
by the subset relation and form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements. The bottom scenario of $\lattice{\system}$ is
$\emptyset$, the top one $\system$ itself; in between are the mixed-typed
scenarios.

@@

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The {\em root\/} $\conf_0$ of a trail is
the initial debugging scenario.  Starting from such a root, the rational
programmer steps along a trail. If the (program for this) scenario type-checks,
the rational programmer runs the program until it raises a run-time error.  The
rational programmer uses the information in the error to decide which module to
equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$
and thus creates a trail. The trail's construction ends when it reaches a
scenario that contains the mutated module because the type checker rejects its
typed version outright.  At this point, the source of the impedance mismatch is
identified.

>>>>

\citet{tfgnvf-popl-2016} describe the set of all possible type migrations with a
lattice. A program $\system$ is a set of modules $\set{\component}$.
Let a configuration $\conf$ of $\system$ be the subset of $\set{\component}$
that comes with type annotations. These configurations of $\system$ are ordered
by the subset relation and form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements. The bottom of $\lattice{\system}$ is
$\emptyset$, the top one $\set{\component}$ itself; in between are the mixed-typed ones.

@@

The actions of the rational programmer create an ascending chain---dubbed a
\emph{trail}---in $\lattice{\system}$.  The {\em root\/} $\conf_0$ of a trail is
the initial debugging scenario. If the program for some scenario  $\conf_i$ type-checks,
the rational programmer runs the program until it raises a run-time error.  The
rational programmer then uses the information in the error to decide which module to
equip with types. This choice constructs scenario $\conf_{i+1}$ from $\conf_{i}$
and thus creates a trail. The trail's construction ends when it reaches a
scenario that contains the mutated module because the type checker rejects its
typed version outright.  At this point, the source of the impedance mismatch is
identified.

### 2021-05-20 writing/rational-transient.tex
tex problems fixed
tighten prose, get fig. 3 on same page as first reference

The Transient semantics assigns blame to a sequence of modules. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame set. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the module that is added to the blame sequence first and
assign types to only that one---after all, if fully checked, the types of this
first module should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
module that is added to the blame sequence last, effectively interpreting the
blame sequence as a boundary-aware stack.

@@

\begin{quote}
\it A \emph{Transient-first blame trail}/\emph{Transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and
\[ \conf_{i+1} \setminus \conf_i = %
\left\{ %
\begin{tabular}{ll} %
$\{\first{\mblame{\system, \conf_i}}\}$  & if $\conf_i$ produces blame \\ %
$\{\exception{\system, \conf_i}\}$     & otherwise \\ %
\end{tabular} %
\]
where $\first{\mblame{\system, \conf}}$/$\last{\mblame{\system, \conf}}$ is the
first/last module, respectively,  that Transient adds to the blame sequence for $\conf$.
\end{quote}

The definition of \emph{Transient exception trails} is analogous to that for
Natural. It is used to as a baseline for Transient-first and Transient-last
blame.

>>>>

The Transient semantics assigns blame to a sequence of modules. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the sequence, and each crossing
checked the value's type in a shallow manner.

This ambiguity in Transient blame raises the question of how the rational
programmer should react when the language produces a blame set. Our answer
is that the rational programmer has at least two reasonable options. The first
one is to select the untyped module that is added to the blame sequence first and
assign types to only that one---after all, if fully checked, the types of this
first module should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to select the
module that is added to the blame sequence last, effectively interpreting the
blame sequence as a boundary-aware stack.

@@

\begin{quote}
\it A \emph{Transient-first blame trail}/\emph{Transient-last blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and
\[ \conf_{i+1} \setminus \conf_i = %
\left\{ %
\begin{tabular}{ll} %
$\{\first{\mblame{\system, \conf_i}}\}$  & if $\conf_i$ produces blame \\ %
$\{\exception{\system, \conf_i}\}$     & otherwise \\ %
\end{tabular} \right. %
\]
where $\first{\mblame{\system, \conf}}$/$\last{\mblame{\system, \conf}}$ is the
first/last module, respectively,  that Transient adds to the blame sequence for $\conf$.
\end{quote}

The definition of \emph{Transient exception trails} is analogous to that for
Natural. It is used to as a baseline for Transient-first and Transient-last
blame.\footnote{A reader may wonder whether the rational programmer
should not just equip {\em all\/} modules in the Transient blame set
with types. Doing so may accelerate the search for the impedance
mismatch but it would also impose a large migration cost.}

### 2021-05-20 writing/rational.tex
tighten prose, get fig. 3 on same page as first reference

Section~\ref{sec:why-rational} explains how a migratory type setting helps with
finding the source of an impedance mismatch. Roughly speaking, it encourages the
rational programmer to equip a module with types if it is blamed in an error
message. A sequence of such steps make up a portion of the migratory path in the
lattice of type migration~\cite{tfgnvf-popl-2016}. The lattice serves as the
common substrate for the definition modes of the rational programmer (see
sec.~\ref{sub:stuff}).

Each mode receives different kinds of information and thus may construct
different paths in the lattice. As discussed in section~\ref{sec:challenges},
evaluating blame relies on comparing modes of the rational programmer
within the same semantics and across different semantics.

Hence the research problem is develop modes for the rational programmer and to make
them comparable even when they correspond to different semantics and offer
different kinds of information (see secs.~\ref{sub:natural}
through~\ref{sub:erasure}). Programmer effort relative to a fixed debugging
scenario introduces another dimension along which the modes become may be
compared (see sec.~\ref{sub:effort}). With these notions in place, it
becomes possible to state the experimental questions and gives an overview of
the experimental process (see sec.~\ref{sub:experiment}).

>>>>

Section~\ref{sec:why-rational} explains how a migratory type setting helps with
finding the source of an impedance mismatch. Roughly speaking, it encourages the
rational programmer to equip a module with types if it is blamed in an error
message. A sequence of such steps makes up a path in the
lattice of type migration~\citep{tfgnvf-popl-2016}. The lattice serves as the
universe \mf{what does this sentence really mean? what does it contribute?} for the definition modes of the rational programmer (see
sec.~\ref{sub:stuff}).

Each mode receives different kinds of information and thus may construct
different paths in the lattice. As discussed in section~\ref{sec:challenges},
evaluating blame relies on comparing modes of the rational programmer
within the same semantics and across different semantics.  
Hence the research problem is to develop modes for the rational programmer and to make
them comparable even when they correspond to different semantics and process
different kinds of information (see secs.~\ref{sub:natural}
through~\ref{sub:erasure}). Programmer effort relative to a fixed debugging
scenario introduces another dimension along which the modes become may be
compared (see sec.~\ref{sub:effort}). With these notions in place, it
becomes possible to state the experimental questions and describes the process (see sec.~\ref{sub:experiment}).
